Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
section 7 completed

Missing values are often ignored when plotting data - which is why the data visualisation software, MANET, was named using an acronym for "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.

@@

The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, with `vis_miss()` from \pkg{visdat} [@visdat]. This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D). Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \@ref(fig:gg-miss-case-var)D). Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation. Using \pkg{ggplot2}, as the foundation, makes the plot easily customizable.

@@

The number of times observations are missing together can be visualized using an "upset plot" [@Conway2017]. An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, and scales well with more variables. An upset plot can be constructed from the shadow matrix, as shown in Figure  \@ref(fig:airquality-upset) which displays the overlapping counts of  missings in the airquality data. The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable. This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables Solar.R and Ozone.

@@

Missing values are typically not shown for univariate visualizations such as histograms or densities. Two ways to use _nabular_ data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below_all`. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the histogram of missing ozone values on the left, in orange. The missings in Ozone are imputed and `Ozone_NA` is  mapped to the fill aesthetic. Nabular data facilitates adding counts of missingness to a histogram, allowing examination of a variables' distribution of values, and also the magnitude of missings.

**Univariate split by missingness**. The distribution of a variable can be shown according to the missingness of another variable. The shadow matrix part of nabular is used to handle the faceting, and color mapping. Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing, a faceted histogram (B), and  an overlaid density (C).  This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values. This type of plot can facilitate exploring missing data distributions. For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.

@@

Parallel coordinate plots can help to visualize missingness beyond two dimensions. They transform variables to the same scale, ranging between 0 and 1. The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds. Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña. Figure \@ref(fig:parallel-cord-plot) is a parallel coordinate plot of `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to whether humidity was missing (`humidity_NA`). Figure \@ref(fig:parallel-cord-plot) shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.

@@

This section shows how the methods described so far are used together in a data analysis workflow. These are applied to a case study, of housing data for the city of Melbourne from January 28, 2016 to March 17, 2018. The data was compiled by scraping weekly property clearance data [@Kaggle-2018-data]. There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset. The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from Central Business District (CBD).

@@

Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values. The most missings are in building area, followed by year built, and land size, with similar amounts of missingness in Car, bathroom, bedroom2, longitude, and latitude. Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing. The variables building area and year built have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious. Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).

@@

Figure \@ref(fig:applic-vis-miss) shows missingness patterns with an `upset` plot [@Conway2017], displaying 8 intersecting sets of missing variables. Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).

@@

Tabulating the number of missings in variables in Table \@ref(tab:housing-miss-var-case-table) (left) shows three groups of missingness. Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right) shows six patterns of missingness. These overview plots lead to the removal of two variables from with more than 50% missingness from analysis: building area and year built.

@@

Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: Land size, latitude, longitude, bedroom2, bathroom, car, and land size.

Missingness structure is explored by clustering the missing values into groups. Then, a  classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables, to find variables and values that predicting missingness clusters [@Tierney2015; @Barnett2017]. Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot].  Importance scores reveal variables important for predicting missingness: rooms, price, suburb, council area, distance, and region name. These variables are important for predicting missingness, so are important to include in the imputation model.

@@

\pkg{simputation} is used to implement two imputation: simple linear regression and K nearest neighbors. Values are imputed stepwise in ascending order of missingness. The  track missings pattern is applied (described in \@ref(verbs-track)), to assess imputed values. Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \@ref(fig:imputed-by-model)). Compared to KNN imputed values, the linear model imputed values closer to the mean.

@@

## Assess model predictions {#case-study-assess-model}

@@

The `naniar` and `visdat` packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex. Each tool has clear intent and effects - plotting or generating data or augmenting data in some way. This reduces repetition and typing for the user, making exploration of missing values easier as they follow consistent, predictable rules.

>>>>

Missing values are often ignored when plotting data - so much so that the data visualisation software, MANET, is an acronym for "Missings Are Now Equally Treated" [@Unwin1996]. However, plots can help to reveal the type of missing value patterns, even those of MCAR, MAR or MNAR. Here we summarise how to systematically explore missing patterns visually, and define useful plots to make, relative to the nabular data structure.

@@

The shadow matrix allows both the variables and cases to be displayed using a heatmap style visualization, `vis_miss()` from \pkg{visdat} [@visdat]. This also provides numerical summaries of missingness in the legend, and for each column (Figure \@ref(fig:gg-miss-case-var)C-D). Clustering can be applied to the rows, and columns arranged in order of missingness (Figure \@ref(fig:gg-miss-case-var)D). Similar visualizations are available in other packages such as \pkg{VIM}, \pkg{mi}, \pkg{Amelia}, and \pkg{MissingDataGUI}. A key improvement is `vis_miss()` orients the visualization analogous to a regular data structure: variables form columns and are named at the top, and each row is an observation. Using \pkg{ggplot2} makes the plot easily customizable.

@@

The number of times observations are missing together is visualized using an "upset plot" [@Conway2017]. An alternative to a Venn diagram, an upset plot shows the size and features of overlapping sets, scaling well with more variables. An upset plot can be constructed from the shadow matrix, with Figure  \@ref(fig:airquality-upset) visualising overlapping amounts missings in the airquality data. The bottom right shows the combinations of missingness, the top panel shows the size of these combinations, and the bottom left shows missingness in each variable. This provides similar information to Figure \@ref(fig:gg-miss-case-var)A-D, but more clearly illustrating overlapping missingness, where 2 cases are missing together in variables Solar.R and Ozone.

@@

Missing values are by default not shown for univariate visualizations such as histograms or densities. Two ways to use _nabular_ data to present univariate data with missings are discussed. The first imputes values below the range to facilitate visualizations. The second displays two plots of the same variable according to the missingness of a chosen variable.

@@

**Imputing values below the range**. To visualize the amount of missings in each variable, the data is transformed into _nabular_ form, then values are imputed below the range of data using `impute_below`. Figure \@ref(fig:impute-shift-histogram)A shows a histogram of Ozone values on the right in green, and the  count of missing ozone values on the left, in orange. The missings in Ozone are mapped to the fill aesthetic in ggplot by referring to the variable `Ozone` as `Ozone_NA`, the values imputed below the range. Nabular data facilitates adding counts of missingness to a histogram, allowing not only understanding of a variables' distribution of values, but also understanding of the magnitude of missings.

**Univariate split by missingness**. Two distributions of a variable can be shown according to the missingness of another variable. Figure \@ref(fig:impute-shift-histogram) shows the values of temperature when ozone is present, and missing. Figure \@ref(fig:impute-shift-histogram)A is a faceted histogram, and Figure \@ref(fig:impute-shift-histogram)B is an overlaid density. This shows how values of temperature are affected by the missingness of ozone, and reveals a cluster of low temperature observations with missing ozone values.
The nabular structure of the data helps reveals the distribution of the missing values. This can facilitate exploring missing data distributions. For example, we would expect if data were MCAR, for values to be roughly uniformly missing thoughout the histogram or density.

@@

Parallel coordinate plots can visualize missingness beyond two dimensions. They transform variables to the same scale, ranging between 0 and 1. The `oceanbuoys` dataset from \pkg{naniar} is used for this visualisation, containing measurements of sea and air temperature, humidity, and east west and north south wind speeds. Data was collected in 1993 and 1997, to understand and predict El Niño and El Niña. Figure \@ref(fig:parallel-cord-plot) is a parallel coordinate plot of `oceanbuoys`, with missing values imputed to be 10% below the range, and values colored according to whether humidity was missing (`humidity_NA`). Figure \@ref(fig:parallel-cord-plot) shows humidity is missing at low air and sea temperatures, and humidity is missing in one year, and one location.

@@

This section shows how the methods described so far are used together in a data analysis workflow. We analyse a case study of housing data for the city of Melbourne from January 28, 2016 to March 17, 2018. The data was compiled by scraping weekly property clearance data [@Kaggle-2018-data]. There are `r scales::number(nrow(housing), big.mark = ",")` properties, and `r ncol(housing)` variables in the dataset. The variables include real estate type (town house, unit, house), suburb, selling method, number of rooms, price, real estate agent, sale date, and distance from Central Business District (CBD).

@@

Figure \@ref(fig:housing-miss-case-var)A shows 9 variables with missing values. The most missings are in "building area", followed by "year built", and "land size", with similar amounts of missingness in "Car", "bathroom", "bedroom2", "longitude", and "latitude." Figure \@ref(fig:housing-miss-case-var)B reveals there are up to 50% missing values in cases, and the majority of cases have more than 5% values missing. The variables "building area" and "year built" have more than 50% missing data, and so could perhaps be omitted from subsequent analysis, as imputed values are likely to be spurious. Three missingness clusters are revealed by visualizing missingness in the whole dataset, clustering and arranging the rows and columns of the data (Figure \@ref(fig:applic-vis-miss)).

@@

Figure \@ref(fig:housing-upset) shows missingness patterns with an `upset` plot [@Conway2017], displaying 8 intersecting sets of missing variables. Two patterns stand out: two, and five variables missing, providing further evidence of the missingness patterns seen in Figures \@ref(fig:housing-miss-case-var) and \@ref(fig:housing-upset).

@@

Tabulating the number of missings in variables in Table \@ref(tab:housing-miss-var-case-table) (left) shows three groups of missingness. Tabulating missings in cases (Table \@ref(tab:housing-miss-var-case-table) (right) shows six patterns of missingness. These overview plots lead to the removal of two variables from with more than 50% missingness from analysis: "building area" and "year built".

@@

Using information from \@ref(case-study-explore-pattern), the following variables are explored for features predicting missingness: "Land size", "latitude", "longitude", "bedroom2", "bathroom", "car", and "land size".

Missingness structure is explored by clustering the missing values into groups. Then, a  classification and regression trees (CART) model is applied to predict these missingness clusters using the remaining variables [@Tierney2015; @Barnett2017]. Two clusters of missingness are identified and predicted using all variables in the dataset with the CART package `rpart` [@rpart], and plotted using the `rpart.plot` package [@rpart-plot]. Importance scores reveal the following variables as most important in predicting missingness: "rooms", "price", "suburb", "council area", "distance", and "region name". These variables are important for predicting missingness, so are included in the imputation model.

@@

\pkg{simputation} is used to implement two imputations: simple linear regression and K nearest neighbors. Values are imputed stepwise in ascending order of missingness. The  track missings pattern is applied (described in \@ref(verbs-track)), to assess imputed values. Imputed datasets are compared on their performance in a model predicting log house price for 4 variables (Figure \@ref(fig:imputed-by-model)). Compared to KNN imputed values, the linear model imputed values closer to the mean.

@@

## Assessing model predictions {#case-study-assess-model}

@@

The `naniar` and `visdat` packages implement the methods discussed in the paper, building on existing tidy tools and strike a compromise between automation and control, making analysis efficient, readable, but not overly complex. Each tool has clear intent and effects - summarising, plotting or generating data or augmenting data in some way. This not only reduces repetition and typing in an analysis, but most importantly, allows for clear expression of intent, making exploration of missing values fluent.
